---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# STMdevelopment

<!-- badges: start -->
<!-- badges: end -->

Develops state-and-transition models for new Ecological Site Groupings in the Upper Colorado River Basin. Inputs include plot data from AIM, NRI, and other projects; soil characteristics; and climate data. Outputs are RMarkdown files formatted like the Ecological Site Description pages in EDIT.

## Installation

You can install the released version of STMdevelopment from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("annack84/STMdevelopment")
```
## Example

No examples yet.

```{r example, eval=FALSE, echo=FALSE}
library(STMdevelopment)
## basic example code
```


```{r cars, eval=FALSE, echo=FALSE}
summary(cars)
```


```{r pressure, echo = FALSE, eval=FALSE}
# You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/master/examples>.

#You can also embed plots, for example:
plot(pressure)
```


