---
title: "Exploring Clustering and Ordination Techniques"
author: "Anna Knight"
date: "10/10/2021"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 1
    toc_float: true
    number_sections: false
---

<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 14px;
}
h1.title {
  font-size: 30px;
}
h1 { /* Header 1 */
  font-size: 26px;
}
h2 { /* Header 2 */
    font-size: 22px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# Helpful resources
  1. "Analysis of Ecological Communities" by McCune & Grace - Great place to start to understand the statistics! Detailed, but not overwhelming. Provides specific recommendations helpful for community ecology. Does NOT provide any R coding help. (Cheapest hardcopies usually from the authors' website: https://www.wildblueberrymedia.net/store/analysis-of-ecological-communities)
  2. "Numerical Ecology" by Legendre & Legendre - Loads of detail about the stats. Comprehensive descriptions of almost every ordination and clustering method available. Recommends some R packages but doesn't provide code. Available free on DOI VPN here: https://www.sciencedirect.com/bookseries/developments-in-environmental-modelling/vol/24/suppl/C
  3. Vegan R package vignette - Lots of handy R coding demos for ordination. Light on stats descriptions. See https://cran.r-project.org/web/packages/vegan/vignettes/intro-vegan.pdf 
  4. Some additional handy online tutorials: https://www.rpubs.com/an-bui/vegan-cheat-sheet and  http://www.umass.edu/landeco/teaching/multivariate/schedule/summary.handouts.pdf
  5. Your pals in the Moab UseR group!

# Community analysis goals and workflow
In this demo, I'm trying to identify plant communities within a set of plots that occur on similar soil types (ask Travis about his Ecological Site Groups if you want the nitty-gritty). I want to know how distinct the plant communities are from one another, and whether there are any major environmental gradients that separate them.

The general workflow is:

  1. Format your data, pick a **distance metric**, and use those to calculate a distance matrix
  2. Group the plots with a **cluster analysis**
  3. Determine the appropriate number of groups (many method options - I'll do an **indicator species analysis**)
  4. Investigate relationships among groups (e.g. do they overlap a lot?) with **ordination**
  5. Look at how environmental variables correspond to the plant community structure with more ordination-related tools
  
I've structured this around plant community analysis because that's my jam, but there are plenty of other reasons to use clustering and ordination - feel free to bring up your own scenarios as we go through this!

# Data prep and distance matrix
```{r setup, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
target_ESG <- "Semiarid_Warm_SandyUplands_LoamyUplands"
user <- "Anna"
#user <- "VPN"

#devtools::install_github("annack84/STMdevelopment")
library(STMdevelopment)
```

```{r visible-setup, include=TRUE}
library(dplyr) # for data wrangling
library(kableExtra) # for formatting tables in RMarkdown
library(vegan) # for clustering and ordination!
library(ggplot2, quietly = T)
```

```{r load-plot-indicators, include=FALSE}
indicators <- c("AH_C3NativePerenGrassCover", 
                "AH_C3IntroducedPerenGrassCover",
                "AH_C4NativePerenGrassCover", 
                "AH_C4IntroducedPerenGrassCover",
                #"AH_IntroducedPerenGrassCover", # Non-native perennial grasses
                "AH_NativePerenForbCover", # Native perennial forbs
                "AH_IntroducedPerenForbCover", # Non-native perennial forbs
                "AH_NativeAnnGrassCover", # Native annual grasses
                "AH_IntroducedAnnGrassCover", # Non-native annual grasses
                "AH_NativeAnnForbCover", # Native annual forbs
                "AH_IntroducedAnnForbCover", # Non-native annual forbs
                "AH_ArtemisiaTridentataCover", # All big sagebrush species lumped together
                "BareSoilCover", # Bare soil
                "CP_percent_100to200", # Canopy gaps > 100 cm TODO do we want annual or perennial gaps?
                "CP_percent_200plus",
                "FH_LichenCover", # Lichen + moss combined cover
                "FH_MossCover",
                "SoilStab_all" # to represent cyano abundance... I think this is only marginally a community variable. If used in hierarchical clustering and Bray-Curtis NMDS, will need to standardize it to make units more similar to % cover units
                )

shrub_by_spp <- T # All shrubs and sub-shrubs by species
subshrub_by_spp <- T
tree_by_spp <- T # All trees by species
opuntia_combined <- T

data_sources <- c("BadgerWash",
                             "CRCMonitoring",
                             "IM_NCPN",
                             "LMF",
                             #"NRI",
                             "Parashant",
                             "AIM", 
                             "VanScoyocThesis"
                           )

plot_data <- plot_data_pull(user=user,
                             target_ESG = target_ESG,
                           data_sources = data_sources,
                           indicators = indicators,
                           shrub_by_spp = shrub_by_spp,
                           subshrub_by_spp = subshrub_by_spp,
                           tree_by_spp = tree_by_spp,
                           opuntia_combined = opuntia_combined
                           )

# remove Artemisia tridentata individual subspecies - use AH_ArtemisiaTridentataCover instead
plot_data <- select(plot_data,
                     -any_of(c("ARTR2",
                               "ARTRW8",
                               "ARTRT",
                               "ARTRV",
                               "ARTRP4",
                               "ARTRV2",
                               "ARTRW",
                               "ARTRS2", 
                               "ARTRX")))


```

Decide what variables to include. Often in community ecology, people include every species as its own variable. This data set covers a large area with a LOT of herbaceous species, so we lumped them into functional groups pertinent to our objectives. 

This report was run on `r Sys.Date()`. The PlotNet files used are those with the most recent dates before `r Sys.Date()`.

```{r doc-plot-indicators, echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
indicator_descriptions <- make_indicator_descriptions(indicators = indicators,
                                                      shrub_by_spp = shrub_by_spp,
                                                      subshrub_by_spp = subshrub_by_spp,
                                                      tree_by_spp = tree_by_spp,
                                                      opuntia_combined = opuntia_combined)

kbl(x = indicator_descriptions[,-1], format = "html", row.names = F, align = "c",
    caption = "Indicators used in ordination and clustering analysis to develop plant communities") %>%
   kable_styling(bootstrap_options = "bordered") %>%
  row_spec(kable_input=., row = 0:nrow(indicator_descriptions), background = "lightsteelblue") %>%
  collapse_rows(columns = 1, valign = "top")
```

### Other data prep things to consider:
  * Sometimes it helps to drop out species/variables with non-zero values in only a few plots (<5% of plots or sample units). NOTE that this can really change your results!
  * For presence/absence data with a lot of absences (zeros), consider converting the data to probability of presence with Beals smoothing
  * Decide whether to relativize the data.
    + Lots of ways to do this (see Ch 9. in McCune & Grace). Relative cover is a common one for plant cover data.
      * relative cover = species cover/total foliar cover
    + Useful if species composition matters to your objectives more than absolute cover (rare species will get more weight)
    + Also helpful if you have variables in different units (e.g. cover and count data)
      * e.g. my soil stability dilemma
  * Decide what to do about incomplete or missing values, if you have them

```{r ordination-data, echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# remove incomplete rows - Bray Curtis distance can't work with NAs
plot_data_clean <- na.omit(plot_data)

# Some plots were sampled multiple times - I'll include only the first year that a plot has complete data
plot_data_first <- plot_data_clean %>%
  dplyr::group_by(PlotCode) %>%
  dplyr::arrange(.data=., Year, .by_group = TRUE) %>%
  dplyr::filter(dplyr::row_number()==1) %>%
  dplyr::ungroup()

# make the clustering and ordination data frame - can't include ANY columns except the actual variables!
ord.df <- dplyr::select(plot_data_first, -SourceKey, -PlotID, -SiteName, -PlotName,
                        -Year, -Longitude_NAD83, -Latitude_NAD83,
                        -Month, -Day) %>%
  tibble::column_to_rownames("PlotCode") # keep an ID code for the plot as the row name to prevent data scrambling problems

# remove rare species
sp_pa <- mutate(rowwise(ord.df), across(everything(), function(x){if(x>0){1}else{0}})) # make a presence/absence data frame
sp_keep <- names(colSums(sp_pa)[which(colSums(sp_pa)>2)])
ord.df <- select(ord.df, all_of(sp_keep))


```

### Distance measures (aka dissimilarity measures)
  * Quantify how different each unit is from another
  * Can calculate among rows (i.e. how different each plot is from another) or columns (i.e. how different each species is from another)
  * LOTS of options here! Some common ones:
    + Euclidean - basic geometry. Uses the Pythagorean theorem on as many dimensions as you have variables
    + Bray-Curtis (aka Sorensen) - One of the most common in community ecology. Works with binary data (e.g. presence/absence) as well as cover, abundance, etc. No negative numbers in your data allowed! 

```{r distance, echo=TRUE}
dist.df <- vegan::vegdist(ord.df, method = "bray") # creates Bray-Curtis distance matrix
```

# Clustering
  * Lots of ways to find groups in data!
    + Hierarchical clustering (many flavors) - groups are nested on branches of a tree
    + K-means clustering - groups are not nested
      * Fuzzy clustering - plots are assigned a probability of being in each group
    + Multivariate classification trees (CART, random forest, etc.) - supervised (you pre-define the groups with training data) or unsupervised (you decide how many groups and then let the machine learning process define them)
    + TWINSPAN

## Hierarchical clustering
There are dozens of ways to do hierarchical clustering! The two things you have to choose are a distance measure (see above) and a linkage method compatible with your distance measure. A common combo that works well for a lot of ecological community data is to use Bray-Curtis distance with a flexible beta linkage (β = –0.25).

```{r clustering, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.cap="Initial clustering dendrogram"}
# cluster and indicator species analysis
library(cluster) # for cluster analysis
library(dendextend) # for pretty cluster plotting

#clust <- hclust(dist.df, method = "average")
clust1 <- agnes(x= dist.df, diss=T, method="flexible", par.method = c(0.625,0.625,-0.25,0)) # flexible beta with beta=-0.25
#plot(clust1)
hclust1 <- as.hclust(clust1) # covert to class hclust for easier plotting
plot(hclust1, labels = FALSE)
```

## How many groups should I have?
The hardest part about clustering is determining the right number of groups!
Some strategies:
  * Silhouette plots (compare dissimilarity between a plot and its group vs. between a plot and everything not in its group)
  * Maximize the correlation between the distance matrix and a binary grouping matrix
  * Indicator species analysis (i.e. species fidelity)
  
```{r silhouette, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.cap="Silhouette optimal number of clusters"}
Si <- numeric(50)
for(k in 2:50){
  sil <- silhouette(x=cutree(hclust1, k=k),
                    dist=dist.df)
  Si[k] <- summary(sil)$avg.width
}

k.best <- which.max(Si)

plot(1:50,
     Si,
     type = "h",
     xlab = "k (number of clusters)",
     ylab = "Average silhouette width")
axis(1,
     k.best,
     paste("Optimum", k.best, sep = "\n"),
     col = "red",
     font = 2,
     col.axis = "red")
points(k.best,
       max(Si),
       pch=16,
       col = "red",
       cex = 1.5)

```
  
```{r matrix-cor, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.cap="Matrix correlation optimal number of clusters"}

grpdist <- function(X){
  require(cluster)
  gr <- as.data.frame(as.factor(X))
  distgr <- daisy(gr, "gower")
  distgr
}

kt <- data.frame(k=1:50, r=0)

for(i in 2:50) {
  gr <- cutree(hclust1, i)
  distgr <- grpdist(gr)
  mt <- cor(dist.df, distgr, method = "pearson")
  kt [i, 2] <- mt
}

k.best <- which.max(kt$r)

plot(kt$k,
     kt$r,
     type = "h",
     xlab = "k (number of clusters)",
     ylab = "Pearson's correlation")

axis(1,
     k.best,
     paste("Optimum", k.best, sep = "\n"),
     col = "red",
     font = 2,
     col.axis = "red")

points(k.best,
       max(kt$r),
       pch = 16,
       col = "red",
       cex = 1.5)

```
  
```{r indicatorsp, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.cap="Mean p-value of indicator species (left) and number of significant species (right, α = 0.05) for 2 to 12 possible plant communities."}
library(labdsv) # indicator species analysis
library(foreach)

# use IndVal from Dufrene and Legendre (1997)
# groupings from hierarchical clustering (grp)

# create vectors of groupings with different numbers of clusters
grps <- data.frame(site=1:nrow(ord.df))
for (i in 2:12) {
  grps <- cbind(grps, cutree(hclust1, k=i))
}
colnames(grps) <- c("site", paste(2:12, "groups", sep="_"))

# calculate IndVal for each species in each number of clusters
listnames <- paste("indvals", 2:12, sep="")

ivlabs_list <- foreach(i=2:12) %do% indval(x=ord.df, clustering =grps[,i], numitr = 1000)

names(ivlabs_list) <- listnames

# Create a data frame with three columns: number of clusters, average p, number of sig ind spp
ivlabs_df <- data.frame(nclusters=2:12, p_mean=NA, nsignif=NA, sumsignif=NA)
for(i in seq(length(ivlabs_list))){
  ivlabs_df[i,"p_mean"] <- mean(ivlabs_list[[i]]$pval, na.rm = T) 
  ivlabs_df[i, "nsignif"] <- length(which(ivlabs_list[[i]]$pval<=0.05))
  ivlabs_df[i,"sumsignif"] <- sum(ivlabs_list[[i]]$indval, na.rm = T)
}

par(mfrow=c(1, 2))
# plot average p values by number of clusters
plot(x=ivlabs_df$nclusters, y=ivlabs_df$p_mean, xlab="Number of clusters", ylab="Mean p-value")
# plot number of significant indicator species by number of clusters
plot(x=ivlabs_df$nclusters, y=ivlabs_df$nsignif, xlab="Number of clusters", ylab="Number of significant species")
par(mfrow=c(1, 1))

k_groups <- 4 # manually change based on plot results
```

The idea here is to minimize the mean p-value and maximize the number of significant species. Based on the results from the three different tests, the optimal number of vegetation groupings is `r k_groups`. The silhouette method and matrix correlation method both selected 4 groups. Only the indicator species analysis suggested including more.

It's a good idea to check your indicator species groupings with a good botanist to make sure the groupings make sense bioligically! E.g. here, pinyon and juniper got grouped together, which definitely makes botanical sense. However, the sagebrush groupings are a bit weird - I need to go back and lump all sagebrush subspecies together.

```{r cut-tree, echo=TRUE, message=FALSE, warning=FALSE}
# create groups
grp <- cutree(hclust1, k=k_groups)

# create data frame that identifies indicator species by group:
ivlab <- indval(x=ord.df, clustering = grp, numitr = 1000)

gr <- ivlab$maxcls 
iv <- ivlab$indcls 
pv <- ivlab$pval 

indvalsummary <- data.frame(group=gr, indval=iv, pvalue=pv) #, freq=fr)
indvalsummary <- indvalsummary[order(indvalsummary$group, -indvalsummary$indval),]
indvalsummary$Code <- rownames(indvalsummary)

# add in readable indicator names
indvalsummary <- left_join(x=indvalsummary,
                           y=select(indicator_descriptions, Indicator_code, Indicator),
                           by=c("Code"="Indicator_code")) %>%
  left_join(x=.,
            y=select(read.csv(data_file_paths(user)$species_list), SpeciesCode, ScientificName),
            by=c("Code"="SpeciesCode")) %>%
  mutate(Indicator = ifelse(test = is.na(Indicator), yes = ScientificName, no = Indicator))

if(opuntia_combined){
  indvalsummary$Indicator[which(indvalsummary$Code=="AH_OpuntiaCover")] <- indicator_descriptions$Indicator[which(indicator_descriptions$Indicator_code=="opuntia_combined")]
}

# add in relative abundance
relab <- as.data.frame(ivlab$relabu)
colnames(relab) <- paste0("RelativeAbund_", 1:ncol(relab))
relab$Code <- rownames(relab)
indvalsummary <- left_join(x=indvalsummary, y=relab, by="Code")

# add in relative frequency
relfr <- as.data.frame(ivlab$relfrq)
colnames(relfr) <- paste0("RelativeFreq_", 1:ncol(relfr)) 
relfr$Code <- rownames(relfr)
indvalsummary <- left_join(x=indvalsummary, y=relfr, by="Code")

# create a table for display where relative abundance and relative frequency are only displayed for the group for which the species is an indicator
indvalsummary_disp <- dplyr::select(indvalsummary, group, indval, pvalue, Indicator)
indvalsummary_disp$RelAbund <- NA
for(g in 1:k_groups){
 indvalsummary_disp[indvalsummary_disp$group==g, "RelAbund"] <- 
   indvalsummary[indvalsummary$group==g, paste0("RelativeAbund_", g)]
}

indvalsummary_disp$RelFreq <- NA
for(g in 1:k_groups){
 indvalsummary_disp[indvalsummary_disp$group==g, "RelFreq"] <- 
   indvalsummary[indvalsummary$group==g, paste0("RelativeFreq_", g)]
}

# order data frame by group, then significance; for output file
#indvalsummary_disp <- indvalsummary_disp[order(indvalsummary_disp$group, indvalsummary_disp$indval), ]
indvalsummary_disp <- indvalsummary_disp %>%
  group_by(group) %>%
  arrange(desc(indval), .by_group=T)

indvalsummary_disp_filt <- filter(indvalsummary_disp, pvalue<=0.05) %>%
  mutate(indval=round(indval, 2),
         pvalue=round(pvalue, 3),
         RelAbund=round(RelAbund, 2),
         RelFreq=round(RelFreq, 2)) %>%
  select(group, Indicator, indval, pvalue, RelAbund, RelFreq) %>%
  setNames(c("Group", "Indicator", "Indicator value", "p-value", "Relative abundance", "Relative frequency"))

kbl(x = indvalsummary_disp_filt, format = "html", row.names = F, align = "c",
    caption = "Indicator species for plant community groups. The indicator value of a species is a combined measure of fidelity (always present in the group) and specificity (exclusive to the group), where higher values indicate a stronger association with the given vegetation group. Relative abundance indicates specificity, where 1 means the species only occurs in plots of the given group. Relative frequency indicates fidelity, where 1 means the species always occurs in plots of the given group. Only species with a p-value less than 0.05 are displayed.") %>%
  kable_styling(bootstrap_options = "bordered") %>%
  row_spec(kable_input=., row = 0:nrow(indvalsummary_disp_filt), background = "lightsteelblue") %>%
  collapse_rows(columns = 1, valign = "top")

```

Going back to our clustering dendrogram, here's what it looks like when we cut the tree to have `r k_groups` groups.

```{r dendrogram, echo=TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.cap="Cluster dendrogram for plant community groups. Boxes represent the groups identified by the indicator species analysis."}
# Cluster dendrogram plot
hclust1_dend <- as.dendrogram(hclust1)

# veg group color palette
pal_veg <- RColorBrewer::brewer.pal(n=k_groups, name = "Set1")

plot(hclust1_dend, leaflab= "none")
rect.dendrogram(hclust1_dend, # put boxes around each group 
                k=k_groups, 
                cluster = grp, 
                border = pal_veg[unique(cutree(hclust1_dend, k=k_groups)[order.dendrogram(hclust1_dend)])], # this bit puts the groupings in the right order - makes it easier to use the same color for a given group in all your plots
                text = as.character(unique(cutree(hclust1_dend, k=k_groups)[order.dendrogram(hclust1_dend)])), 
                text_cex = 2, 
                lower_rect = -0.25,
                lwd =2)

#plot(ape::as.phylo(hclust1), tip.color = pal_veg[grp], direction = "downwards")
```

```{r silhouette-results, echo=TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.cap="Silhouette plot of the final groupings."}

sil <- silhouette(cutree(hclust1, k=k_groups), dist.df)

rownames(sil) <- rownames(ord.df)
plot(sil,
     cex.names = 0.8,
     col = 2:(k_groups + 1),
     nmax = 100)

```

# Ordination
## Why do an ordination?
  * Data reduction - simplify a complex web of intercorrelated variables 
    + Can create a smaller number of composite variables, e.g. combining many similar climate variables (PCA is great for this!)
  * Visualize patterns
  * Identify strong vs. weak patterns

## Ordination methods
  * Constrained ordination - e.g. canonical correspondence analysis
    + Use with two sets of many variables (explanatory and response)
  * Unconstrained ordination - e.g. principal components analysis (PCA), nonmetric multi-dimensional scaling (NMDS)
    + Use with one set of many variables (e.g. plant cover by species)
    + Can fit another set of variables (e.g. environmental conditions) _after_ running the ordination
    + PCA is best for normal data with roughly linear relationships between variables
    + NMDS works well for non-normal and zero-inflated data (most plant community data!)

Here, we're using ordination to visualize patterns in the plant communities. The data are zero-inflated (more rare species than common species - this is typical) so NMDS is a good choice.

## How many axes?
  * Stress increases as the number of axes decreases
    + Stress below 0.20 is recommended - higher than that and results aren't very reliable
  * For interpretablility, you're really choosing between 2 and 3 axes
  
```{r ordination-scree, echo=TRUE, error=FALSE, warning=FALSE, message=FALSE, results='hide', fig.keep='all', cache=FALSE, fig.cap="NMDS ordination scree plot comparing number of dimensions in the solution to stress."}
# This part takes a while!
NMDS_scree(ord.df)
ord_dims <- 3 # change manually based on the plot results
```

```{r ordination, eval=TRUE, include=TRUE}

#run NMS ordination 
set.seed(1) # always set a seed if you want to be able to rerun your code and get the exact same ordination results!
ord <- metaMDS(ord.df,
               k=ord_dims, # number of dimensions
               trymax = 30)  # can increase this number if you're having convergence problems

```

Based on Figure \@ref(fig:ordination-scree), ordination stress is acceptable with `r ord_dims` dimensions, and stress reduction is minimal with additional dimensions. The stress for the ordination with `r ord_dims` dimensions is `r round(ord$stress, 3)`..

```{r ordination-plots, echo=TRUE, fig.height=10, fig.width=10, fig.cap="NMDS ordination of plot indicator data. Points represent plots; ellipses represent plant communities identified by the hierarchical clustering and indicator species analysis."}
# 2D plotting
plot_data_first$PlantCommunity <- as.factor(grp)

# manually update based on indicator species
# group_labels <- c("1 - Sagebrush shrubland",
#                     "2 - Wyoming big sagebrush shrubland",
#                     "3 - Sparse shrubland",
#                     "4 - C3 perennial grassland with
# mixed shrubs",
#                     "5 - Pinyon-juniper woodland",
#                     "6 - Mixed oak shrubland",
#                     "7 - Introduced annual herbaceous 
# community",
#                     "8 - Ephedra shrubland with biocrust",
#                     "6 - C4 perennial grassland"
# )

group_labels <- c("1",
                  "2",
                  "3",
                  "4")

par(mfrow=c(2,2))
# Axes 1x2
plot(ord, choices = c(1,2), type = "n", # plot the axes
     xlim = c(-1.25, 1.25),
     ylim = c(-1, 1))
points(ord, choices = c(1,2), display = "sites", # plot points - can choose "sites" or "species"
       col=pal_veg[plot_data_first$PlantCommunity],
       pch = 21, cex = .6, bg=pal_veg[plot_data_first$PlantCommunity])
ordiellipse(ord, plot_data_first$PlantCommunity, col=pal_veg, lwd = 2, label = T,
            choices = c(1,2)) # plot your groups
# can use oriellipse, orihull, or ordispider to plot groups, depending on your needs

# Axes 3x2
plot(ord, choices = c(3,2), type = "n", 
     xlim = c(-1.25, 1.25),
     ylim = c(-1, 1))
points(ord, choices = c(3,2), display = "sites", 
       col=pal_veg[plot_data_first$PlantCommunity],
       pch = 21, cex = .6, bg=pal_veg[plot_data_first$PlantCommunity])
ordiellipse(ord, plot_data_first$PlantCommunity, col=pal_veg, lwd = 2, label = T,
            choices = c(3,2)) 

# Axes 1x3
plot(ord, choices = c(1,3), type = "n",
     xlim = c(-1.25, 1.25),
     ylim = c(-1, 1))
points(ord, choices = c(1,3), display = "sites",
       col=pal_veg[plot_data_first$PlantCommunity],
       pch = 21, cex = .6, bg=pal_veg[plot_data_first$PlantCommunity])
ordiellipse(ord, plot_data_first$PlantCommunity, col=pal_veg, lwd = 2, label = T,
            choices = c(1,3)) 

# Legend
plot(ord, type = "n", axes=FALSE,
     display = "sites",
     col=pal_veg[plot_data_first$PlantCommunity],
     xlab = "",
     ylab = "")
legend(x="center",
       legend = group_labels,
       fill = pal_veg,
       title = "Plant communities")

par(mfrow=c(1,1))

```

```{r ordination-spplots, echo=TRUE, fig.height=10, fig.width=10, fig.cap="NMDS ordination of plot indicator data with plant species overlayed."}


# Axes 1x2
plot(ord, choices = c(1,2), type = "n", # plot the axes
     xlim = c(-1.25, 1.25),
     ylim = c(-1, 1))
text(ord, choices = c(1,2), display = "species")
ordiellipse(ord, plot_data_first$PlantCommunity, col=pal_veg, lwd = 2, label = T) # plot your groups


```

You can plot in 3 dimensions if you're feeling fancy!

```{r threeD-plots, eval=TRUE, echo=TRUE, fig.height=5, fig.width=10, fig.cap="Surface diagrams for environmental variables"}
# make a 3d plot
library(vegan3d)

invisible(rgl::open3d())

ordirgl(ord, display = "sites", type = "n", alpha = 1,
        arr.col = "blue"
        ) # alpha is transparency of color - 0 is fully transparent, 1 is not transparent
orglellipse(ord, groups = plot_data_first$PlantCommunity, kind = "sd", col = pal_veg, alpha = 0.4)

rgl::rglwidget() # this embeds the 3D plot in the HTML doc (and it's zoomable! Ooooooooh, aaaaaaaah!)
```

## Fitting environmental variables
To examine relationships between the plant community structure and environmental variables, you can fit an environmental matrix to your ordination.

The standard arrow diagrams assume linear relationships. Surface plots can help you decide if the relationship is approximately linear.

I didn't have time to put together the environmental data for this demo, but here's some code to try it out on your own!

```{r env-plots, eval=FALSE, echo=TRUE, fig.height=5, fig.width=10, fig.cap="Arrow diagrams for environmental variables"}
ord.full <- read.csv("path_to_some_environmental_data")

ord.fit <- envfit(ord ~ Elevation_ft,
                  data = ord.full,
                  na.rm=T,
                  choices=c(1:3)) # "choices" indicates number of dimensions

par(mfrow=c(1,2))
# Axes 1x2
plot(ord, choices = c(1,2), type = "n", # plot the axes. "choices" indicates which pair of axes to plot.
     xlim = c(-1.25, 1.25),
     ylim = c(-1, 1))
points(ord, choices = c(1,2), display = "sites", # plot points - can choose "sites" or "species"
       col=pal_veg[plot_data_first$PlantCommunity],
       pch = 21, cex = .6, bg=pal_veg[plot_data_first$PlantCommunity])
ordiellipse(ord, plot_data_first$PlantCommunity, col=pal_veg, lwd = 2, label = T) # plot your groups
plot(ord.fit, add=T)

# Legend
plot(ord, type = "n", axes=FALSE,
     display = "sites",
     col=pal_veg[plot_data_first$PlantCommunity],
     xlab = "",
     ylab = "")
legend(x="center",
       legend = group_labels,
       fill = pal_veg,
       title = "Plant communities")

par(mfrow=c(1,1))

```

```{r env-plots2, eval=FALSE, echo=TRUE, fig.height=5, fig.width=10, fig.cap="Surface diagrams for environmental variables"}
plot(ord, choices=c(1,3))
ordisurf(ord~Elevation_ft, ord.full, labcex=1, choices=c(1,3)) # non-linear
plot(ord.fit_Elev, choices=c(1,3))
```

# Fuzzy clustering
Fuzzy c-means clustering uses Euclidean distances and least-squares solutions to find the optimal grouping for each site. This means that our data going into the clustering algorithm need to be normalized. Borcard, Gillet, and Legendre (2011) suggest two ways of doing this. One is to used the "normalize" in the function vegan::decostand() on the raw species matrix. The other is to take the square root of the Bray-Curtis distance matrix, run a principal coordinates analysis (PCoA), take the site scores from the PCoA, and run them through the clustering algorithm (I think? see pp. 97-98 in Borcard et al. 2011).

The first option sound simpler, so let's try that:

```{r normalize, echo=TRUE}
norm.df <- vegan::decostand(x=ord.df, method = "normalize")
chord.df <- vegdist(x=norm.df, method = "euclidean")
# NOTE: normalizing and then computing Euclidean distance is called "chord distance".
```

## What is the optimal number of fuzzy groups?

We'll test splitting into 2 to 12 fuzzy groups. The best solution will maximize both the normalized Dunn's partition coefficient and the average silhouette width.

```{r num-fuzzy-grps, echo=TRUE, fig.width=10, fig.cap="Normalized Dunn's index and average silhouette width for 2 to 12 possible groups in fuzzy clustering."}
# test out groupings of 2 to 12
res<- data.frame()
for (i in 2:12) {super=fanny(chord.df, #d.bray,
                             k=i,
                             diss=T,
                             #maxit=500,
                             #metric = "euclidean",
                             memb.exp= 1.2 #1.1
)
res=rbind(res, c(i, super$convergence[1], super$coeff[2], super$silinfo$avg.width))
colnames(res)<-c("i", "inter", "dunn.index.normalized", "silhouette.averaged.width")
}

k.best.dunn <- res$i[which.max(res$dunn.index.normalized)]
k.best.sil <- res$i[which.max(res$silhouette.averaged.width)]

par(mfrow= c(1,2))
plot(x=res$i,
     y=res$dunn.index.normalized,
     type = "h",
     xlab = "k (number of clusters)",
     ylab = "Normalized Dunn's partition coef.")
axis(1,
     k.best.dunn,
     paste("Optimum", k.best.dunn, sep = "\n"),
     col = "red",
     font = 2,
     col.axis = "red")
points(k.best.dunn,
       max(res$dunn.index.normalized),
       pch = 16,
       col = "red",
       cex = 1.5)

plot(x=res$i,
     y=res$silhouette.averaged.width,
     type = "h",
     xlab = "k (number of clusters)",
     ylab = "Silhouette averaged width")
axis(1,
     k.best.sil,
     paste("Optimum", k.best.sil, sep = "\n"),
     col = "red",
     font = 2,
     col.axis = "red")
points(k.best.sil,
       max(res$silhouette.averaged.width),
       pch = 16,
       col = "red",
       cex = 1.5)

par(mfrow= c(1,1))

```


```{r fuzzy-cluster, echo=TRUE}
k_groups_fuzzy <- 3
clust_fuzz <- fanny(x=chord.df,
                    k = k_groups_fuzzy,
                    diss = T,
                    #metric = "euclidean",
                    memb.exp = 1.2 # this affects how crisp/fuzzy the clusters are - closer to 1 is crisper
                    )

```

```{r sil-plot-custom, echo=TRUE, fig.height=5, fig.width=5, fig.cap="Silhouette width by group."}
# veg group color palette
pal_veg2 <- RColorBrewer::brewer.pal(n=k_groups_fuzzy, name = "Dark2")

sil_fuzzy <- as.data.frame(clust_fuzz$silinfo$widths)
sil_fuzzy$cluster <- as.factor(as.character(sil_fuzzy$cluster))

boxplot(sil_width~cluster,
        data = sil_fuzzy,
        ylab = "Silhouette width",
        notch = T,
        col = pal_veg2
        )

```

## Ordination of the fuzzy clusters
Borcard et al. (2011) recommend using PCoA ordination to visualized fuzzy clustering results.

```{r pcoa, echo=TRUE, fig.height=10, fig.width=10, fig.cap="Ordination of fuzzy clusters (PCoA"}
dc.pcoa <- cmdscale(chord.df, k=3, eig = T, add = T) # if add = T, you have applied a Cailliez correction (helps reduce problems with negative eigenvalues)


par(mfrow=c(2,2))
choice_list <- list(c(1,2), c(3,2), c(1,3))
for(c in 1:length(choice_list)){
  choices = choice_list[[c]]
  
  dc.scores <- scores(dc.pcoa, choices = choices)

plot(dc.scores,
     asp = 1,
     type = "n",
     )

abline(h=0, lty = "dotted")
abline(v=0, lty = "dotted")

stars(
  clust_fuzz$membership,
  labels = NULL,
  location = dc.scores,
  key.loc = c(0.6, 0.6),
  key.labels = 1:k_groups_fuzzy,
  draw.segments = T,
  add = T,
  len = 0.075,
  col.segments = pal_veg2[1:(k_groups_fuzzy)]
)

for(i in 1:k_groups_fuzzy){
  gg <- dc.scores[clust_fuzz$clustering == i, ]
  hpts <- chull(gg)
  hpts <- c(hpts, hpts[1])
  lines(gg[hpts, ], col = pal_veg2[i], lwd = 2)
}

}
par(mfrow=c(1,1))

```

```{r pca, eval=FALSE, echo=FALSE, fig.height=5, fig.width=10, fig.cap="Ordination of fuzzy clusters (PCA"}
spe.h.pca <- rda(norm.df)

screeplot(spe.h.pca,
          bstick = T,
          npcs = length(spe.h.pca$CA$eig))
# okay so don't do PCA, it wants me to interpret 8 axes!
```

```{r ordination-scree2, echo=TRUE, error=FALSE, warning=FALSE, message=FALSE, results='hide', fig.keep='all', cache=FALSE, fig.cap="NMDS ordination scree plot comparing number of dimensions in the solution to stress."}
# This part takes a while!
NMDS_scree_euc<-function(x) { #x is the name of the data frame
  plot(rep(1,2),
       replicate(2,vegan::metaMDS(x,k=1, distance = "euclidean")$stress),
       xlim=c(1,6),
       ylim=c(0,1),
       xlab="# of Dimensions",
       ylab="Stress",
       main="NMDS stress plot")
  for (i in 1:3) {
    points(rep(i+1,2),
           replicate(2,vegan::metaMDS(x, distance = "euclidean",k=i+1)$stress))
  }}

NMDS_scree_euc(norm.df)
ord_dims2 <- 3 # change manually based on the plot results
```

```{r ordination2, eval=TRUE, include=TRUE}

#run NMS ordination 
set.seed(1) # always set a seed if you want to be able to rerun your code and get the exact same ordination results!
ord2 <- metaMDS(norm.df,
               k=ord_dims2, # number of dimensions
               trymax = 50, # can increase this number if you're having convergence problems
               distance = "euclidean"
               )  

```

Based on Figure \@ref(fig:ordination-scree2), ordination stress is acceptable with `r ord_dimss` dimensions, and stress reduction is minimal with additional dimensions. The stress for the ordination with `r ord_dims2` dimensions is `r round(ord2$stress, 3)`..

```{r ordination-plots2, echo=TRUE, fig.height=10, fig.width=10, fig.cap="NMDS ordination of normalized plot indicator data. Points represent plots; ellipses represent plant communities identified by fuzzy clustering analysis."}
# 2D plotting
grp_fuzz <- clust_fuzz$clustering
plot_data_first$PlantCommunity_fuzzy <- as.factor(grp_fuzz)

# manually update based on indicator species
# group_labels <- c("1 - Sagebrush shrubland",
#                     "2 - Wyoming big sagebrush shrubland",
#                     "3 - Sparse shrubland",
#                     "4 - C3 perennial grassland with
# mixed shrubs",
#                     "5 - Pinyon-juniper woodland",
#                     "6 - Mixed oak shrubland",
#                     "7 - Introduced annual herbaceous 
# community",
#                     "8 - Ephedra shrubland with biocrust",
#                     "6 - C4 perennial grassland"
# )

group_labels <- c("1",
                  "2",
                  "3")

par(mfrow=c(2,2))
# Axes 1x2
plot(ord2, choices = c(1,2), type = "n", # plot the axes
     xlim = c(-0.75, 0.75),
     ylim = c(-0.75, 0.75))
points(ord2, choices = c(1,2), display = "sites", # plot points - can choose "sites" or "species"
       col=pal_veg2[plot_data_first$PlantCommunity_fuzzy],
       pch = 21, cex = .6, bg=pal_veg2[plot_data_first$PlantCommunity_fuzzy])
ordiellipse(ord2, plot_data_first$PlantCommunity_fuzzy, col=pal_veg2, lwd = 2, label = T,
            choices = c(1,2)) # plot your groups
# can use oriellipse, orihull, or ordispider to plot groups, depending on your needs

# Axes 3x2
plot(ord2, choices = c(3,2), type = "n", 
     xlim = c(-0.75, 0.75),
     ylim = c(-0.75, 0.75))
points(ord2, choices = c(3,2), display = "sites", 
       col=pal_veg2[plot_data_first$PlantCommunity_fuzzy],
       pch = 21, cex = .6, bg=pal_veg2[plot_data_first$PlantCommunity_fuzzy])
ordiellipse(ord2, plot_data_first$PlantCommunity_fuzzy, col=pal_veg2, lwd = 2, label = T,
            choices = c(3,2)) 

# Axes 1x3
plot(ord2, choices = c(1,3), type = "n",
     xlim = c(-0.75, 0.75),
     ylim = c(-0.75, 0.75))
points(ord2, choices = c(1,3), display = "sites",
       col=pal_veg2[plot_data_first$PlantCommunity_fuzzy],
       pch = 21, cex = .6, bg=pal_veg2[plot_data_first$PlantCommunity_fuzzy])
ordiellipse(ord2, plot_data_first$PlantCommunity_fuzzy, col=pal_veg2, lwd = 2, label = T,
            choices = c(1,3)) 

# Legend
plot(ord2, type = "n", axes=FALSE,
     display = "sites",
     col=pal_veg2[plot_data_first$PlantCommunity_fuzzy],
     xlab = "",
     ylab = "")
legend(x="center",
       legend = group_labels,
       fill = pal_veg2,
       title = "Plant communities")

par(mfrow=c(1,1))

```

Other things to try:
- put the star plot in physical space to look for geographic patterns
- What to do with the fuzziest plots (lowest silhouette width, and/or group membership below some threshold)
- Indicator species analysis on the fuzzy cluster results? To help describe the communities, and provide indicator species for end-users
- plot plant functional type gradients on whichever ordination you decide to go with (also helps describe communities)


```{r echo=FALSE}
# TO MAKE THE 3D PLOT WORK, DON'T USE THE KNIT BUTTON! COPY AND PASTE THE CODE BELOW INTO THE CONSOLE!!
#rmarkdown::render(input = "C:/Users/aknight/Desktop/Telework_Backups/V_drive/PROJECTS/ESG/STM/Scripts/STMdevelopment/R/ordination_clustering_demo.Rmd", output_format = "html_document", output_file = "C:/Users/aknight/Desktop/Telework_Backups/V_drive/PROJECTS/ESG/STM/ordination_clustering_demo.html")
```
