---
title: "Exploring Clustering and Ordination Techniques"
author: "Anna Knight"
date: "9/30/2021"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 1
    toc_float: true
    number_sections: false
---

<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 14px;
}
h1.title {
  font-size: 30px;
}
h1 { /* Header 1 */
  font-size: 26px;
}
h2 { /* Header 2 */
    font-size: 22px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# Helpful resources
  1. "Analysis of Ecological Communities" by McCune & Grace - Great place to start to understand the statistics! Detailed, but not overwhelming. Provides specific recommendations helpful for community ecology. Does NOT provide any R coding help. (Cheapest hardcopies usually from the authors' website: https://www.wildblueberrymedia.net/store/analysis-of-ecological-communities)
  2. "Numerical Ecology" by Legendre & Legendre - Loads of detail about the stats. Comprehensive descriptions of almost every ordination and clustering method available. Recommends some R packages but doesn't provide code. Available free on DOI VPN here: https://www.sciencedirect.com/bookseries/developments-in-environmental-modelling/vol/24/suppl/C
  3. Vegan R package vignette - Lots of handy R coding demos for ordination. Light on stats descriptions. See https://cran.r-project.org/web/packages/vegan/vignettes/intro-vegan.pdf 
  4. Some additional handy online tutorials: https://www.rpubs.com/an-bui/vegan-cheat-sheet and  http://www.umass.edu/landeco/teaching/multivariate/schedule/summary.handouts.pdf
  5. Your pals in the Moab UseR group!

# Community analysis goals and workflow
In this demo, I'm trying to identify plant communities within a set of plots that occur on similar soil types (ask Travis about his Ecological Site Groups if you want the nitty-gritty). I want to know how distinct the plant communities are from one another, and whether there are any major environmental gradients that separate them.

The general workflow is:
  1. Format your data, pick a **distance metric**, and use those to calculate a distance matrix
  2. Group the plots with a **cluster analysis**
  3. Determine the appropriate number of groups (many method options - I'll do an **indicator species analysis**)
  4. Investigate relationships among groups (e.g. do they overlap a lot?) with **ordination**
  5. Look at how environmental variables correspond to the plant community structure with more ordination-related tools
  
I've structured this around plant community analysis because that's my jam, but there are plenty of other reasons to use clustering and ordination - feel free to bring up your own scenarios as we go through this!

# Data prep and distance matrix
```{r setup, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
target_ESG <- "Semiarid_Warm_SandyUplands_LoamyUplands"
user <- "Anna"
#user <- "VPN"


library(STMdevelopment)
```

```{r visible-setup, include=TRUE}
library(dplyr) # for data wrangling
library(kableExtra) # for formatting tables in RMarkdown
library(vegan) # for clustering and ordination!
```

```{r load-plot-indicators, include=FALSE}
indicators <- c("AH_C3NativePerenGrassCover", 
                "AH_C3IntroducedPerenGrassCover",
                "AH_C4NativePerenGrassCover", 
                "AH_C4IntroducedPerenGrassCover",
                "AH_IntroducedPerenGrassCover", # Non-native perennial grasses
                "AH_NativePerenForbCover", # Native perennial forbs
                "AH_IntroducedPerenForbCover", # Non-native perennial forbs
                "AH_NativeAnnGrassCover", # Native annual grasses
                "AH_IntroducedAnnGrassCover", # Non-native annual grasses
                "AH_NativeAnnForbCover", # Native annual forbs
                "AH_IntroducedAnnForbCover", # Non-native annual forbs
                "BareSoilCover", # Bare soil
                "CP_percent_100to200", # Canopy gaps > 100 cm TODO do we want annual or perennial gaps?
                "CP_percent_200plus",
                "FH_LichenCover", # Lichen + moss combined cover
                "FH_MossCover")

shrub_by_spp <- T # All shrubs and sub-shrubs by species
subshrub_by_spp <- T
tree_by_spp <- T # All trees by species
opuntia_combined <- T

data_sources <- c("BadgerWash",
                             "CRCMonitoring",
                             "IM_NCPN",
                             "LMF",
                             #"NRI",
                             "Parashant",
                             "AIM", 
                             "VanScoyocThesis"
                           )

plot_data <- plot_data_pull(user=user,
                             target_ESG = target_ESG,
                           data_sources = data_sources,
                           indicators = indicators,
                           shrub_by_spp = shrub_by_spp,
                           subshrub_by_spp = subshrub_by_spp,
                           tree_by_spp = tree_by_spp,
                           opuntia_combined = opuntia_combined
                           )


```

Decide what variables to include. Often in community ecology, people include every species as its own variable. This data set covers a large area with a LOT of herbaceous species, so we lumped them into functional groups pertinent to our objectives. 

```{r doc-plot-indicators, echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
indicator_descriptions <- make_indicator_descriptions(indicators = indicators,
                                                      shrub_by_spp = shrub_by_spp,
                                                      subshrub_by_spp = subshrub_by_spp,
                                                      tree_by_spp = tree_by_spp,
                                                      opuntia_combined = opuntia_combined)

kbl(x = indicator_descriptions[,-1], format = "html", row.names = F, align = "c",
    caption = "Indicators used in ordination and clustering analysis to develop plant communities") %>%
   kable_styling(bootstrap_options = "bordered") %>%
  row_spec(kable_input=., row = 0:nrow(indicator_descriptions), background = "lightsteelblue") %>%
  collapse_rows(columns = 1, valign = "top")
```

### Other data prep things to consider:
  * Sometimes it helps to drop out species/variables with non-zero values in only a few plots (<5% of plots or sample units).
  * For presence/absence data with a lot of absences (zeros), consider converting the data to probability of presence with Beals smoothing
  * Decide whether to relativize the data.
    + Lots of ways to do this (see Ch 9. in McCune & Grace). Relative cover is a common one for plant cover data.
      * relative cover = species cover/total foliar cover
    + Useful if species composition matters to your objectives more than absolute cover (rare species will get more weight)
    + Also helpful if you have variables in different units (e.g. cover and count data)
      * e.g. my soil stability dilemma
  * Decide what to do about incomplete or missing values, if you have them

```{r ordination-data, echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# remove incomplete rows - Bray Curtis distance can't work with NAs
plot_data_clean <- na.omit(plot_data)

# Some plots were sampled multiple times - I'll include only the first year that a plot has complete data
plot_data_first <- plot_data_clean %>%
  dplyr::group_by(PlotCode) %>%
  dplyr::arrange(.data=., Year, .by_group = TRUE) %>%
  dplyr::filter(dplyr::row_number()==1) %>%
  dplyr::ungroup()

# make the clustering and ordination data frame - can't include ANY columns except the actual variables!
ord.df <- dplyr::select(plot_data_first, -SourceKey, -PlotID, -SiteName, -PlotName,
                        -Year, -Longitude_NAD83, -Latitude_NAD83,
                        -Month, -Day) %>%
  tibble::column_to_rownames("PlotCode") # keep an ID code for the plot as the row name to prevent data scrambling problems

# Keep only species that occur in at least 2 plots
ord.df <- ord.df[ ,which(colSums(ord.df[,1:ncol(ord.df)] !=0) >= 2)] 

head(ord.df)

```

### Distance measures (aka dissimilarity measures)
  * Quantify how different each unit is from another
  * Can calculate among rows (i.e. how different each plot is from another) or columns (i.e. how different each species is from another)
  * LOTS of options here! Some common ones:
    + Euclidean - basic geometry. Uses the Pythagorean theorem on as many dimensions as you have variables
    + Bray-Curtis (aka Sorensen) - One of the most common in community ecology. Works with binary data (e.g. presence/absence) as well as cover, abundance, etc. No negative numbers in your data allowed! 

```{r distance, echo=TRUE}
dist.df <- vegan::vegdist(ord.df, method = "bray") # creates Bray-Curtis distance matrix
```

# Clustering
  * Lots of ways to find groups in data!
    + Hierarchical clustering (many flavors) - groups are nested on branches of a tree
    + K-means clustering - groups are not nested
      * Fuzzy clustering - plots are assigned a probability of being in each group
    + Multivariate classification trees (CART, random forest, etc.) - supervised (you pre-define the groups with training data) or unsupervised (you decide how many groups and then let the machine learning process define them)
    + TWINSPAN

## Hierarchical clustering
There are dozens of ways to do hierarchical clustering! The two things you have to choose are a distance measure (see above) and a linkage method compatible with your distance measure. A common combo that works well for a lot of ecological community data is to use Bray-Curtis distance with a flexible beta linkage (β = –0.25).

```{r clustering, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.cap="Initial clustering dendrogram"}
# cluster and indicator species analysis
library(cluster) # for cluster analysis
library(dendextend) # for pretty cluster plotting

#clust <- hclust(dist.df, method = "average")
clust1 <- agnes(x= dist.df, diss=T, method="flexible", par.method = c(0.625,0.625,-0.25,0)) # flexible beta with beta=-0.25
#plot(clust1)
hclust1 <- as.hclust(clust1) # covert to class hclust for easier plotting
plot(hclust1, labels = FALSE)
```

## How many groups should I have?
The hardest part about clustering is determining the right number of groups!
Some strategies:
  * Silhouette plots (compare dissimilarity between a plot and its group vs. between a plot and everything not in its group)
  * Maximize the correlation between the distance matrix and a binary grouping matrix
  * Indicator species analysis (i.e. species fidelity)
  
```{r indicatorsp, echo=TRUE, fig.cap="Mean, fig.width=10, message=FALSE, warning=FALSE, fig.cap="Mean p-value of indicator species (left) and number of significant species (right, α = 0.05) for 2 to 12 possible plant communities."}
library(labdsv)
library(foreach)
# indicator species
# use IndVal from Dufrene and Legendre (1997)
# groupings from hierarchical clustering (grp)

# create vectors of groupings with different numbers of clusters
grps <- data.frame(site=1:nrow(ord.df))
for (i in 2:12) {
  grps <- cbind(grps, cutree(hclust1, k=i))
}
colnames(grps) <- c("site", paste(2:12, "groups", sep="_"))

# calculate IndVal for each species in each number of clusters
listnames <- paste("indvals", 2:12, sep="")

ivlabs_list <- foreach(i=2:12) %do% indval(x=ord.df, clustering =grps[,i], numitr = 1000)

names(ivlabs_list) <- listnames

# Create a data frame with three columns: number of clusters, average p, number of sig ind spp
ivlabs_df <- data.frame(nclusters=2:12, p_mean=NA, nsignif=NA, sumsignif=NA)
for(i in seq(length(ivlabs_list))){
  ivlabs_df[i,"p_mean"] <- mean(ivlabs_list[[i]]$pval, na.rm = T) 
  ivlabs_df[i, "nsignif"] <- length(which(ivlabs_list[[i]]$pval<=0.05))
  ivlabs_df[i,"sumsignif"] <- sum(ivlabs_list[[i]]$indval, na.rm = T)
}

par(mfrow=c(1, 2))
# plot average p values by number of clusters
plot(x=ivlabs_df$nclusters, y=ivlabs_df$p_mean, xlab="Number of clusters", ylab="Mean p-value")
# plot number of significant indicator species by number of clusters
plot(x=ivlabs_df$nclusters, y=ivlabs_df$nsignif, xlab="Number of clusters", ylab="Number of significant species")
par(mfrow=c(1, 1))

k_groups <- 8 # manually change based on plot results
```

The idea here is to minimize the mean p-value and maximize the number of significant species. Based on the results in Figure \@ref(fig:clustering), the optimal number of vegetation groupings is `r k_groups` (mean p-value = `r round(ivlabs_df$p_mean[ivlabs_df$nclusters==k_groups], 3)`; number of indicator species = `r ivlabs_df$nsignif[ivlabs_df$nclusters==k_groups]`).

```{r cut-tree, echo=TRUE, message=FALSE, warning=FALSE}
# create groups
grp <- cutree(hclust1, k=k_groups)

# create data frame that identifies indicator species by group:
ivlab <- indval(x=ord.df, clustering = grp, numitr = 1000)

gr <- ivlab$maxcls 
iv <- ivlab$indcls 
pv <- ivlab$pval 

indvalsummary <- data.frame(group=gr, indval=iv, pvalue=pv) #, freq=fr)
indvalsummary <- indvalsummary[order(indvalsummary$group, -indvalsummary$indval),]
indvalsummary$Code <- rownames(indvalsummary)

# add in readable indicator names
indvalsummary <- left_join(x=indvalsummary,
                           y=select(indicator_descriptions, Indicator_code, Indicator),
                           by=c("Code"="Indicator_code")) %>%
  left_join(x=.,
            y=select(read.csv(data_file_paths(user)$species_list), SpeciesCode, ScientificName),
            by=c("Code"="SpeciesCode")) %>%
  mutate(Indicator = ifelse(test = is.na(Indicator), yes = ScientificName, no = Indicator))

if(opuntia_combined){
  indvalsummary$Indicator[which(indvalsummary$Code=="AH_OpuntiaCover")] <- indicator_descriptions$Indicator[which(indicator_descriptions$Indicator_code=="opuntia_combined")]
}

# add in relative abundance
relab <- as.data.frame(ivlab$relabu)
colnames(relab) <- paste0("RelativeAbund_", 1:ncol(relab))
relab$Code <- rownames(relab)
indvalsummary <- left_join(x=indvalsummary, y=relab, by="Code")

# add in relative frequency
relfr <- as.data.frame(ivlab$relfrq)
colnames(relfr) <- paste0("RelativeFreq_", 1:ncol(relfr)) 
relfr$Code <- rownames(relfr)
indvalsummary <- left_join(x=indvalsummary, y=relfr, by="Code")

# create a table for display where relative abundance and relative frequency are only displayed for the group for which the species is an indicator
indvalsummary_disp <- dplyr::select(indvalsummary, group, indval, pvalue, Indicator)
indvalsummary_disp$RelAbund <- NA
for(g in 1:k_groups){
 indvalsummary_disp[indvalsummary_disp$group==g, "RelAbund"] <- 
   indvalsummary[indvalsummary$group==g, paste0("RelativeAbund_", g)]
}

indvalsummary_disp$RelFreq <- NA
for(g in 1:k_groups){
 indvalsummary_disp[indvalsummary_disp$group==g, "RelFreq"] <- 
   indvalsummary[indvalsummary$group==g, paste0("RelativeFreq_", g)]
}

# order data frame by group, then significance; for output file
#indvalsummary_disp <- indvalsummary_disp[order(indvalsummary_disp$group, indvalsummary_disp$indval), ]
indvalsummary_disp <- indvalsummary_disp %>%
  group_by(group) %>%
  arrange(desc(indval), .by_group=T)

indvalsummary_disp_filt <- filter(indvalsummary_disp, pvalue<=0.05) %>%
  mutate(indval=round(indval, 2),
         pvalue=round(pvalue, 3),
         RelAbund=round(RelAbund, 2),
         RelFreq=round(RelFreq, 2)) %>%
  select(group, Indicator, indval, pvalue, RelAbund, RelFreq) %>%
  setNames(c("Group", "Indicator", "Indicator value", "p-value", "Relative abundance", "Relative frequency"))

kbl(x = indvalsummary_disp_filt, format = "html", row.names = F, align = "c",
    caption = "Indicator species for plant community groups. The indicator value of a species is a combined measure of fidelity (always present in the group) and specificity (exclusive to the group), where higher values indicate a stronger association with the given vegetation group. Relative abundance indicates specificity, where 1 means the species only occurs in plots of the given group. Relative frequency indicates fidelity, where 1 means the species always occurs in plots of the given group. Only species with a p-value less than 0.05 are displayed.") %>%
  kable_styling(bootstrap_options = "bordered") %>%
  row_spec(kable_input=., row = 0:nrow(indvalsummary_disp_filt), background = "lightsteelblue") %>%
  collapse_rows(columns = 1, valign = "top")

```

Going back to our clustering dendrogram, here's what it looks like with o
```{r dendrogram, echo=TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.cap="Cluster dendrogram for plant community groups. Boxes represent the groups identified by the indicator species analysis."}
# Cluster dendrogram plot
hclust1_dend <- as.dendrogram(hclust1)

# veg group color palette
pal_veg <- RColorBrewer::brewer.pal(n=k_groups, name = "Set1")

plot(hclust1_dend, leaflab= "none")
rect.dendrogram(hclust1_dend, 
                k=k_groups, 
                cluster = grp, 
                border = pal_veg[unique(cutree(hclust1_dend, k=k_groups)[order.dendrogram(hclust1_dend)])], 
                text = as.character(unique(cutree(hclust1_dend, k=k_groups)[order.dendrogram(hclust1_dend)])), 
                text_cex = 2, 
                lower_rect = -0.25,
                lwd =2)

#plot(ape::as.phylo(hclust1), tip.color = pal_veg9[grp], direction = "downwards")
```

# Ordination
## Why do an ordination?
  * Data reduction - simplify a complex web of intercorrelated variables 
    + Can create a smaller number of composite variables, e.g. combining many similar climate variables (PCA is great for this!)
  * Visualize patterns
  * Identify strong vs. weak patterns

## Ordination methods
  * Constrained ordination - e.g. canonical correspondence analysis
    + Use with two sets of many variables (explanatory and response)
  * Unconstrained ordination - e.g. principal components analysis (PCA), nonmetric multi-dimensional scaling (NMDS)
    + Use with one set of many variables (e.g. plant cover by species)
    + Can fit another set of variables (e.g. environmental conditions) _after_ running the ordination
    + PCA is best for normal data with roughly linear relationships between variables
    + NMDS works well for non-normal and zero-inflated data (most plant community data!)

```{r ordination-scree, echo=TRUE, error=FALSE, warning=FALSE, message=FALSE, results='hide', fig.keep='all', cache=FALSE, fig.cap="NMDS ordination scree plot comparing number of dimensions in the solution to stress."}
NMDS_scree(ord.df)
ord_dims <- 2 # change manually based on the plot results
```

```{r ordination, eval=TRUE, include=TRUE}

#run NMS ordination 
set.seed(1)
ord <- metaMDS(ord.df,
               k=ord_dims, # number of dimensions
               trymax = 30) 

```

Based on Figure \@ref(fig:ordination-scree), ordination stress is acceptable with `r ord_dims` dimensions, and stress reduction is minimal with additional dimensions. The stress for the ordination with `r ord_dims` dimensions is `r round(ord$stress, 3)`.

```{r ordination-plots, echo=TRUE, fig.height=5, fig.width=10, fig.cap="NMDS ordination of plot indicator data. Points represent plots; ellipses represent plant communities identified by the hierarchical clustering and indicator species analysis."}
# 2D plotting
plot_data_first$PlantCommunity <- as.factor(grp)

# manually update based on indicator species
group_labels <- c("1 - Perennial grassland with
mixed shrubs and forbs",
                    "2 - Juniper and shrub community
with non-native grasses",
                    "3 - Mixed grassland and shrub community",
                    "4 - Sparse shrub community")

par(mfrow=c(1,2))
# Axes 1x2
plot(ord, choices = c(1,2), type = "n", 
     xlim = c(-1.25, 1.25),
     ylim = c(-1, 1))
points(ord, choices = c(1,2), display = "sites",
       col=pal_veg[plot_data_first$PlantCommunity],
       pch = 21, cex = .6, bg=pal_veg[plot_data_first$PlantCommunity])
ordiellipse(ord, plot_data_first$PlantCommunity, col=pal_veg, lwd = 2, label = T)

# Legend
plot(ord, type = "n", axes=FALSE,
     display = "sites",
     col=pal_veg[plot_data_first$PlantCommunity],
     xlab = "",
     ylab = "")
legend(x="center",
       legend = group_labels,
       fill = pal_veg,
       title = "Plant communities")

par(mfrow=c(1,1))

```
