---
title: 'Ecological Site Group Description: Semiarid Warm Sandy & Loamy Uplands'
author: "Travis Nauman and Anna Knight"
date: "3/10/2021"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 1
    toc_float: true
    number_sections: false
---

<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 14px;
}
h1.title {
  font-size: 30px;
}
h1 { /* Header 1 */
  font-size: 26px;
}
h2 { /* Header 2 */
    font-size: 22px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# General information
```{r setup, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
target_ESG <- "Semiarid_Warm_SandyUplands_LoamyUplands"
user <- "Travis"

library(dplyr)
library(kableExtra)
library(STMdevelopment)
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO insert dot graphic and text about the status of the ESD (e.g. "Provisional"). Might help
# to create a function to automatically insert the description based on the
# status input
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 1. Mapped extent"}
# TODO create leaflet map of the ESG extent
# EDIT_map(target_ESG = target_ESG, user = user)
```

Areas shown in blue indicate the maximum mapped extent of this ecological site. Other ecological sites likely occur within the highlighted areas. It is also possible for this ecological site to occur outside of highlighted areas if detailed soil survey has not been completed or recently updated.

## Associated ecological site concepts

This ecological site group includes a variety of upland ecological sites that have deeper sandy soils without significant rock content. The group was defined by placing ecological sites into groups by separating soils, climate, and geomorphic features to best differentiate reference vegetation production values and documented ecological states as documented ny Nauman et al., (In Review).

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
esds <- esd_data_pull(user)
comps <- comp_data_pull(user)
num_esds <- nrow(esds)
```
This ecological site group includes `r num_esds` ecological sites. These sites were correlated to the group based on average soil, climate and geomorphic properties of the soil components linked to each ecological site.
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
knitr::kable(x = esds[,c("ecoclassid","ecoclassname")], col.names=c("ESD Code","ESD Name"), caption = NULL,row.names = F)
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of dominant plant functional groups and species


# TODO Travis update the esd pull to include top species from reference production??
# knitr::kable(x = dom_plant_spp_tab, caption = "Table 1. Dominant plant species")
```


TODO insert "Download full description" link to a PDF


# Physiographic features

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of landforms, elevations, slopes, water table depth, flooding duration, flooding frequency, ponding frequency
# ideally should be able to tab between US and metric system measurements
landforms <- unlist(strsplit(esds$landfrms[1], split=","))
for(e in 2:nrow(esds)){
  esdlfs <- unlist(strsplit(esds[e,c("landfrms")],split=","))
  landforms <- append(landforms,esdlfs)
}
landforms <- landforms[!landforms %in% "NA"]
lfsum <-  factor(landforms)
lfsum <- reorder(lfsum,lfsum,FUN=function(x) -length(x))
landforms_unique <- unique(landforms)
```
These sites occur on upland positions with mostly gentle slopes on a variety of landforms (n=`r length(landforms_unique)`). Some areas can have steeper slopes, but these are not typical of the group. 

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Slope
slope_ave <- mean(comps@site$slope_r.x)
slope_75th <- unname(quantile(comps@site$slope_r.x, probs = 0.75, na.rm = T))
slope_max <- max(comps@site$slope_r.x, na.rm = T)
slope_sd <- sd(comps@site$slope_r.x, na.rm = T)
slope_field <- paste("Slopes average",format(round(slope_ave, 1), nsmall = 1), 
                     "% with a standard deviation of",format(round(slope_sd, 1), nsmall = 1), 
                     "and are generally less than", format(round(slope_75th, 1), nsmall = 1),
                     "but can be up to", format(round(slope_max, 1), nsmall = 1),"%",sep=" ")

phys.df <- dplyr::tribble(~Item, ~Description,
                          "Landforms (Top 10)", as.character(lfsum)[1:10],
                          "Slope", slope_field,
                          "Flooding Frequency","None",
                          "Aspect", "Aspect is not a significant factor")

knitr::kable(x = phys.df, caption = "Table 2. Representative physiographic features")
```

# Climatic features

Climate is generally semiarid and warm with aridity index (precipitation / potential evapotranspiration) values averaging `r format(round(mean(esds$aimean), 2), nsmall = 2)` and ranging from `r format(round(min(esds$aimean), 2), nsmall = 2)` to `r format(round(max(esds$aimean), 2), nsmall = 2)`. The average maximum temperatures (Celsius) of the hottest month range from `r format(round(min(esds$maxtempmean), 2), nsmall = 2)` to `r format(round(max(esds$maxtempmean), 2), nsmall = 2)`, and minimum temperatures of the coldest month range from `r format(round(min(esds$mintempmean), 2), nsmall = 2)` to `r format(round(max(esds$mintempmean), 2), nsmall = 2)`. Warm season (June to September) precipitation makes up `r format(round(mean(esds$pptrt*100), 0), nsmall = 0)`% of total precipitation on average, but can range from `r format(round(min(esds$pptrt*100), 0), nsmall = 0)`% to `r format(round(max(esds$pptrt*100), 0), nsmall = 0)`%.

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}

clim.df <- dplyr::tribble(~Item, ~Description,
                          "Frost Free Period (days)", paste(format(round(min(comps@site$ffd_r), 0), nsmall = 0),format(round(mean(comps@site$ffd_r), 0), nsmall = 0),format(round(max(comps@site$ffd_r), 0), nsmall = 0),sep=", "),
                          "Mean Annual Precipitation (in)", paste(format(round(min(comps@site$reannualprecip_r/25.4,na.rm=T), 1), nsmall = 1),format(round(mean(comps@site$reannualprecip_r/25.4,na.rm=T), 1), nsmall = 1),format(round(max(comps@site$reannualprecip_r/25.4,na.rm=T), 1), nsmall = 1),sep=", "))

knitr::kable(x = clim.df, caption = "Table 3. Representative climatic features (min, mean, max)")
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 1. Montly precipitation range"}
# TODO create line chart and bar charts of monthly high/low precip that you can tab between
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 1. Montly average minimum and maximum temperature"}
# TODO create line chart and bar charts of monthly high/low temps that you can tab between
```

# Water features

These sites neither benefit significantly from run-in moisture nor experience excessive loss of moisture from runoff.

# Soil features

TODO narrative describing soil features including depth, parent materials, textures, etc.
Soils in this group are moderately deep or deeper to bedrock and are composed primarily of alluvium and eolian sediments. Surface horizons have sand, loamy sand, sandy loam, and loam textures and subsurface horizons are similar but can also include sandy clay loams. Soils are non-saline and non sodic and can have up to 20% calcium carbonate, but generally have less than 6% carbonates. Soil pH ranges from 5.0 to 9.6, but is generally closer to 8.0. Water erosion hazard is moderate and wind erosion hazard is severe.  

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}

## Prep soil variables for table
surftext <-  factor(esds$txtnm_surf)
surftext <- reorder(surftext,surftext,FUN=function(x) -length(x))
surftext_uni <- levels(surftext)
subtext <-  factor(esds$txtnm_sub)
subtext <- reorder(subtext,subtext,FUN=function(x) -length(x))
subtext_uni <- levels(subtext)
drainage <- factor(comps@site$drainagecl.x)
drainage <- reorder(drainage,drainage,FUN=function(x) -length(x))
drainage_uni <- levels(drainage)[1:3]

## Make Table of soil descriptors: 
# TODO Could update pH to be based on esd averages instead of comps. Also could update PM to grab ssurgo values
soil.df <- dplyr::tribble(~Item, ~Description,
                          "Parent Material", "alluvium and eolian sediments",
                          "Surface Texture (0-30cm)", surftext_uni,
                          "Subsurface Texture (>30cm)",subtext_uni,
                          "Drainage", drainage_uni,
                          "Soil Depth", paste(format(round(min(esds$depthmean,na.rm=T), 0), nsmall = 0),"-",format(round(max(esds$depthmean,na.rm=T), 0), nsmall = 0), " cm",sep=""),
                          "Surface Rock Content %vol (0-30cm)", paste(format(round(min(esds$rock_surf,na.rm=T), 0), nsmall = 0),"-",format(round(max(esds$rock_surf,na.rm=T), 0), nsmall = 0), "%",sep=""),
                          "Subsurface Rock Content %vol (>30cm)", paste(format(round(min(esds$rock_sub,na.rm=T), 0), nsmall = 0),"-",format(round(max(esds$rock_sub,na.rm=T), 0), nsmall = 0), "%",sep=""),
                          "Surface Electrical Conductivity (0-30cm)", paste(format(round(min(esds$ec_surf,na.rm=T), 0), nsmall = 0),"-",format(round(max(esds$ec_surf,na.rm=T), 1), nsmall = 1), " dS/m",sep=""),
                          "Subsurface Electrical Conductivity (>30cm)", paste(format(round(min(esds$ec_sub,na.rm=T), 0), nsmall = 0),"-",format(round(max(esds$ec_sub,na.rm=T), 1), nsmall = 1), " dS/m",sep=""),
                          "Surface Sodium Adsorption Ratio (0-30cm)", paste(format(round(min(esds$sar_surf,na.rm=T), 0), nsmall = 0),"-",format(round(max(esds$sar_surf,na.rm=T), 1), nsmall = 1),sep=""),
                          "Subsurface Sodium Adsorption Ratio (>30cm)", paste(format(round(min(esds$sar_sub,na.rm=T), 0), nsmall = 0),"-",format(round(max(esds$sar_sub,na.rm=T), 1), nsmall = 1),sep=""),
                          "Surface 1:1 pH (0-30cm)", paste(format(round(min(comps@horizons[comps@horizons$hzdept_r<30,c("ph1to1h2o_r")],na.rm=T), 0), nsmall = 0),"-",format(round(max(comps@horizons[comps@horizons$hzdept_r<30,c("ph1to1h2o_r")],na.rm=T), 1), nsmall = 1),sep=""),
                          "Subsurface 1:1 pH (>30cm)", paste(format(round(min(comps@horizons[comps@horizons$hzdept_r>30,c("ph1to1h2o_r")],na.rm=T), 0), nsmall = 0),"-",format(round(max(comps@horizons[comps@horizons$hzdept_r>30,c("ph1to1h2o_r")],na.rm=T), 1), nsmall = 1),sep=""),
                          "Surface Calcium Carbonate (0-30cm)", paste(format(round(min(esds$caco3_surf,na.rm=T), 0), nsmall = 0),"-",format(round(max(esds$caco3_surf,na.rm=T), 0), nsmall = 0), "%",sep=""),
                          "Subsurface Calcium Carbonate (>30cm)", paste(format(round(min(esds$caco3_sub,na.rm=T), 0), nsmall = 0),"-",format(round(max(esds$caco3_sub,na.rm=T), 0), nsmall = 0), "%",sep=""))

knitr::kable(x = soil.df, caption = "Table 4. Representative soil features")
```

## Soil Component data

Soil types correlated to this group include `r length(unique(comps@site$cokey))` different components mapped in SSURGO.Soil taxonomic units include:

`r print(unique(comps@site$compname))`

### Component level soil property depth plots
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
agg <- aqp::slab(comps, fm= ~ claytotal_r + silttotal_r + sandtotal_r + ec_r + caco3_r)

lattice::xyplot(top ~ p.q50 | variable, data=agg, ylab='Depth (cm)',
             xlab='median bounded by 25th and 75th percentiles',
             lower=agg$p.q25, upper=agg$p.q75, ylim=c(201,-2),
             panel=aqp::panel.depth_function,
             alpha=0.25, sync.colors=TRUE,
             par.settings=list(superpose.line=list(col='RoyalBlue', lwd=2)),
             prepanel=aqp::prepanel.depth_function,
             cf=agg$contributing_fraction, cf.col='black', cf.interval=20, 
             layout=c(5,1), strip=strip.custom(bg=grey(0.8)),
             scales=list(x=list(tick.number=4, alternating=3, relation='free'))
             )

```

# Ecological dynamics

TODO narrative about key types of transitions and disturbances

## State and transition model {.tabset} 

```{r apriori-stm, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO pull in table of Ecological Site STMs for this ESG
apriori_stms <- read.csv(data_file_paths(user)$apriori_stms, stringsAsFactors = F,
                         na.strings = c("NA", "", " "), header = F)

apriori_stms_target <- dplyr::filter(apriori_stms, V1 %in% c(NA, "ESG", target_ESG))
apriori_stms_target <-  t(apriori_stms_target)[4:41,]
colnames(apriori_stms_target) <- c("State", "Plant community", "Prp. of ESDs containing this state")
# apriori_stms_target <- apriori_stms_target[c(1,2,41,3:40), ] %>%
#   as.data.frame() %>%
#   mutate(V3 = stringr::str_replace_all(V3, Climate_group_definitions_replace)) %>%
#   mutate(V3 = stringr::str_replace_all(V3, SGU_definitions_replace))

kbl(x = apriori_stms_target, format = "html", row.names = F, align = "c",
    caption = "STM summary from Ecological Sites within this ESG") %>%
  kable_styling(bootstrap_options = "bordered") %>%
  row_spec(kable_input=., row = 0:nrow(apriori_stms_target), background = "lightsteelblue") %>%
  collapse_rows(columns = 1, valign = "top")
```

## Reference state production indicators
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
## Plot of reference production indicators used in paper
resp_vars <- colnames(esds)[1:21]
resp_vars <- resp_vars[!resp_vars %in% c("ecoclassid","ponderosa","aspen")]
par(mar = c(4,8,4,2)) ## This works when knited
boxplot(esds[,resp_vars], horizontal = TRUE,las = 1, main = "Reference Production (lbs/Acre)")

```

```{r load-plot-indicators, include=FALSE}
indicators <- c("AH_C3PerenGrassCover", # C3 native perennial grasses TODO calculate C3 NATIVE
                "AH_C4PerenGrassCover", # C4 native perennial grasses TODO calculate C4 NATIVE
                "AH_IntroducedPerenGrassCover", # Non-native perennial grasses
                "AH_NativePerenForbCover", # Native perennial forbs
                "AH_IntroducedPerenForbCover", # Non-native perennial forbs
                "AH_NativeAnnGrassCover", # Native annual grasses
                "AH_IntroducedAnnGrassCover", # Non-native annual grasses
                "AH_NativeAnnForbCover", # Native annual forbs
                "AH_IntroducedAnnForbCover", # Non-native annual forbs
                "BareSoilCover", # Bare soil
                "CP_percent_100to200", # Canopy gaps > 100 cm TODO do we want annual or perennial gaps?
                "CP_percent_200plus",
                "FH_LichenCover", # Lichen + moss combined cover
                "FH_MossCover")

shrub_by_spp <- T # All shrubs and sub-shrubs by species
subshrub_by_spp <- T
tree_by_spp <- T # All trees by species
opuntia_combined <- T

plot_data <- plot_data_pull(user=user,
                             target_ESG = target_ESG,
                           data_sources = c("BadgerWash",
                             "CRCMonitoring",
                             "IM_NCPN",
                             "LMF",
                             "NRI",
                             "Parashant",
                             "TerrADat", 
                             "VanScoyocThesis"
                           ),
                           indicators = indicators,
                           shrub_by_spp = shrub_by_spp,
                           subshrub_by_spp = subshrub_by_spp,
                           tree_by_spp = tree_by_spp,
                           opuntia_combined = opuntia_combined
                           )


```

```{r doc-plot-indicators, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
indicator_descriptions <- make_indicator_descriptions(indicators = indicators,
                                                      shrub_by_spp = shrub_by_spp,
                                                      subshrub_by_spp = subshrub_by_spp,
                                                      tree_by_spp = tree_by_spp,
                                                      opuntia_combined = opuntia_combined)

kbl(x = indicator_descriptions[,-1], format = "html", row.names = F, align = "c",
    caption = "Indicators used to develop plant communities") %>%
  kable_styling(bootstrap_options = "bordered") %>%
  row_spec(kable_input=., row = 0:nrow(indicator_descriptions), background = "lightsteelblue") %>%
  collapse_rows(columns = 1, valign = "top")
```

### CUSTOM DIAGRAM
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="STM"}
# TODO create custom diagram of states
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Legend"}
# TODO legend for custom diagram if needed
```

### STANDARD DIAGRAM

Click on state and transition labels to scroll to the respective text.

TODO create boxes for each state that link to the state description below

## State 1
## Reference State

TODO write state description

## Community 1.1
## Reference State

TODO write state description

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Annual production by plant type (representative values) or group (midpoint values)"}
# TODO create pie charts of community composition by functional group. May need to change "production" to "percent cover" to fit with our data
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of production (or percent cover) by functional type

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 5. Annual production by plant type")
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of ground cover by type

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 6. Ground cover")
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of func type percent cover at different heights

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 7. Canopy structure (% cover)")
```

### Additional community tables

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of production (or percent cover) by species, with species linking to their USDA Plants profile

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 8. Community 1.1 plant community composition")
```

# Interpretations

## Animal community

TODO write wildlife narrative

## Hydrological functions

TODO write hydrology narrative

## Recreational uses

TODO write recreation narrative

## Wood products

TODO write wood production narrative (probably "None")

## Other products

TODO grazing narrative could go here

# Supporting information

## Other references

## Contributors

+ Travis Nauman
+ Samuel Burch
+ Anna Knight
+ Mike Duniway

# Reference sheet

[Interpreting Indicators of Rangeland Health](https://wiki.landscapetoolbox.org/doku.php/field_methods:rangeland_health_assessment_i.e._indicators_of_rangeland_health) is a qualitative assessment protocol used to determine ecosystem condition based on benchmark characteristics described in the Reference Sheet. A suite of 17 (or more) indicators are typically considered in an assessment. The ecological site(s) representative of an assessment location must be known prior to applying the protocol and must be verified based on soils and climate. Current plant community cannot be used to identify the ecological site.

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table with author, author contact info, date, approver, approved date, and "Composition (Indicators 10 and 12) based on" info for the IIRH reference sheet

# knitr::kable(x = dom_plant_spp_tab)
```

## Indicators

TODO fill in the 17 IIRH indicators for the reference state if available

TODO link to PDF of the reference sheet

# Print options

TODO can we actually include this tab in the RMarkdown HTML format?


