---
title: 'Ecological Site Group Description: Semiarid Warm Sandy & Loamy Uplands'
author: "Travis Nauman and Anna Knight"
date: "3/10/2021"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 1
    toc_float: true
    number_sections: false
---

<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 14px;
}
h1.title {
  font-size: 30px;
}
h1 { /* Header 1 */
  font-size: 26px;
}
h2 { /* Header 2 */
    font-size: 22px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# General information
```{r setup, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
target_ESG <- "Semiarid_Warm_SandyUplands_LoamyUplands"
#user <- "Anna"
user <- "VPN"

library(dplyr)
library(kableExtra)
library(STMdevelopment)
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO insert dot graphic and text about the status of the ESD (e.g. "Provisional"). Might help
# to create a function to automatically insert the description based on the
# status input
```

```{r map, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Mapped extent", fig.width=5, fig.height=5}
EDIT_map(target_ESG = target_ESG, user = user)
```

Areas shown in blue indicate the maximum mapped extent of this ecological site. Other ecological sites likely occur within the highlighted areas. It is also possible for this ecological site to occur outside of highlighted areas if detailed soil survey has not been completed or recently updated.

## Associated sites
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of associated sites
# knitr::kable(x = assoc_sites_tab, caption = NULL)
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of dominant plant functional groups and species

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 1. Dominant plant species")
```


TODO insert "Download full description" link to a PDF


# Physiographic features

TODO write a few sentences on the slopes, elevations, and geomorphic characteristics typical of this ESG

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of landforms, elevations, slopes, water table depth, flooding duration, flooding frequency, ponding frequency
# ideally should be able to tab between US and metric system measurements

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 2. Representative physiographic features")
```

# Climatic features

TODO narrative of general climate patterns and growing season goes here (can probably write a basic narrative for most ESGs and insert specific numbers/dates with R chunks)

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of frost free period, freeze-free period, and total precip

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 3. Representative climatic features")
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 1. Montly precipitation range"}
# TODO create line chart and bar charts of monthly high/low precip that you can tab between
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 1. Montly average minimum and maximum temperature"}
# TODO create line chart and bar charts of monthly high/low temps that you can tab between
```

# Water features

# Soil features

TODO narrative describing soil features including depth, parent materials, textures, etc.

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table surface texture, drainage class, permeability class, soil depth class, family particle size, surface and subsurface fragments, AWC, calcium carbonate equivalent, EC, sodium adsorption ratio, soil reaction

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 4. Representative soil features")
```

# Ecological dynamics

TODO narrative about key types of transitions and disturbances

## State and transition model development 

```{r apriori-stm, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO pull in table of Ecological Site STMs for this ESG
apriori_stms <- read.csv(data_file_paths(user)$apriori_stms, stringsAsFactors = F,
                         na.strings = c("NA", "", " "), header = F)

apriori_stms_target <- dplyr::filter(apriori_stms, V1 %in% c(NA, "ESG", target_ESG))
apriori_stms_target <-  t(apriori_stms_target)[4:41,]
colnames(apriori_stms_target) <- c("State", "Plant community", "Proportion of ESDs containing this state")
# apriori_stms_target <- apriori_stms_target[c(1,2,41,3:40), ] %>%
#   as.data.frame() %>%
#   mutate(V3 = stringr::str_replace_all(V3, Climate_group_definitions_replace)) %>%
#   mutate(V3 = stringr::str_replace_all(V3, SGU_definitions_replace))

kbl(x = apriori_stms_target, format = "html", row.names = F, align = "c",
    caption = "STM summary from Ecological Sites within this ESG") %>%
  kable_styling(bootstrap_options = "bordered") %>%
  row_spec(kable_input=., row = 0:nrow(apriori_stms_target), background = "lightsteelblue") %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r load-plot-indicators, include=FALSE}
indicators <- c("AH_C3PerenGrassCover", # C3 native perennial grasses TODO calculate C3 NATIVE
                "AH_C4PerenGrassCover", # C4 native perennial grasses TODO calculate C4 NATIVE
                "AH_IntroducedPerenGrassCover", # Non-native perennial grasses
                "AH_NativePerenForbCover", # Native perennial forbs
                "AH_IntroducedPerenForbCover", # Non-native perennial forbs
                "AH_NativeAnnGrassCover", # Native annual grasses
                "AH_IntroducedAnnGrassCover", # Non-native annual grasses
                "AH_NativeAnnForbCover", # Native annual forbs
                "AH_IntroducedAnnForbCover", # Non-native annual forbs
                "BareSoilCover", # Bare soil
                "CP_percent_100to200", # Canopy gaps > 100 cm TODO do we want annual or perennial gaps?
                "CP_percent_200plus",
                "FH_LichenCover", # Lichen + moss combined cover
                "FH_MossCover")

shrub_by_spp <- T # All shrubs and sub-shrubs by species
subshrub_by_spp <- T
tree_by_spp <- T # All trees by species
opuntia_combined <- T

plot_data <- plot_data_pull(user=user,
                             target_ESG = target_ESG,
                           data_sources = c("BadgerWash",
                             "CRCMonitoring",
                             "IM_NCPN",
                             "LMF",
                             "NRI",
                             "Parashant",
                             "TerrADat", 
                             "VanScoyocThesis"
                           ),
                           indicators = indicators,
                           shrub_by_spp = shrub_by_spp,
                           subshrub_by_spp = subshrub_by_spp,
                           tree_by_spp = tree_by_spp,
                           opuntia_combined = opuntia_combined
                           )


```

```{r doc-plot-indicators, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
indicator_descriptions <- make_indicator_descriptions(indicators = indicators,
                                                      shrub_by_spp = shrub_by_spp,
                                                      subshrub_by_spp = subshrub_by_spp,
                                                      tree_by_spp = tree_by_spp,
                                                      opuntia_combined = opuntia_combined)

kbl(x = indicator_descriptions[,-1], format = "html", row.names = F, align = "c",
    caption = "Indicators used in ordination and clustering analysis to develop plant communities") %>%
  kable_styling(bootstrap_options = "bordered") %>%
  row_spec(kable_input=., row = 0:nrow(indicator_descriptions), background = "lightsteelblue") %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r ordination-data, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# remove incomplete rows - Bray Curtis distance can't work with NAs
plot_data_clean <- na.omit(plot_data)

# include only the first year that a plot has complete data
plot_data_first <- plot_data_clean %>%
  dplyr::group_by(PlotCode) %>%
  dplyr::arrange(.data=., Year, .by_group = TRUE) %>%
  dplyr::filter(dplyr::row_number()==1) %>%
  dplyr::ungroup()

# TODO update I&M species codes to match current USDA Plants codes! probably a step for the plot_data_pull function?

# make the ordination data frame
ord.df <- dplyr::select(plot_data_first, -SourceKey, -PlotID, -SiteName, -PlotName,
                        -Year, -Longitude_NAD83, -Latitude_NAD83,
                        -Month, -Day) %>%
  tibble::column_to_rownames("PlotCode") # keep plot code as row name

# Keep only species that occur in at least 2 plots
ord.df <- ord.df[ ,which(colSums(ord.df[,1:ncol(ord.df)] !=0) >= 2)] 

```

```{r clustering, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.cap="Mean p-value of indicator species (left) and number of significant species (right, α = 0.05) for 2 to 12 possible plant communities."}
# cluster and indicator species analysis
library(cluster) # for cluster analysis
library(dendextend)
library(labdsv)
library(foreach)
library(vegan)

# initial clustering
dist.df <- vegan::vegdist(ord.df, method = "bray") # creates Bray-Curtis distance matrix
#clust <- hclust(dist.df, method = "average") # use agnes() in cluster package for flexible beta method
clust1 <- agnes(x= dist.df, diss=T, method="flexible", par.method = c(0.625,0.625,-0.25,0)) # flexible beta with beta=-0.25
#plot(clust1)
hclust1 <- as.hclust(clust1) # covert to class hclust for easier plotting
#plot(hclust1, labels = FALSE)

# indicator species
# use IndVal from Dufrene and Legendre (1997)
# groupings from hierarchical clustering (grp)
# ALSO CONSIDER USING MRPP

# create vectors of groupings with different numbers of clusters
grps <- data.frame(site=1:nrow(ord.df))
for (i in 2:12) {
  grps <- cbind(grps, cutree(hclust1, k=i))
}
colnames(grps) <- c("site", paste(2:12, "groups", sep="_"))

# calculate IndVal for each species in each number of clusters
listnames <- paste("indvals", 2:12, sep="")

ivlabs_list <- foreach(i=2:12) %do% indval(x=ord.df, clustering =grps[,i], numitr = 1000)

names(ivlabs_list) <- listnames

# Create a data frame with three columns: number of clusters, average p, number of sig ind spp
ivlabs_df <- data.frame(nclusters=2:12, p_mean=NA, nsignif=NA, sumsignif=NA)
for(i in seq(length(ivlabs_list))){
  ivlabs_df[i,"p_mean"] <- mean(ivlabs_list[[i]]$pval, na.rm = T) 
  ivlabs_df[i, "nsignif"] <- length(which(ivlabs_list[[i]]$pval<=0.05))
  ivlabs_df[i,"sumsignif"] <- sum(ivlabs_list[[i]]$indval, na.rm = T)
}

par(mfrow=c(1, 2))
# plot average p values by number of clusters
plot(x=ivlabs_df$nclusters, y=ivlabs_df$p_mean, xlab="Number of clusters", ylab="Mean p-value")
# plot number of significant indicator species by number of clusters
plot(x=ivlabs_df$nclusters, y=ivlabs_df$nsignif, xlab="Number of clusters", ylab="Number of significant species")
par(mfrow=c(1, 1))

k_groups <- 4 # manually change based on plot results
```

Based on the results in Figure \@ref(fig:clustering), the optimal number of vegetation groupings is `r k_groups` (mean p-value = `r round(ivlabs_df$p_mean[ivlabs_df$nclusters==k_groups], 3)`; number of indicator species = `r ivlabs_df$nsignif[ivlabs_df$nclusters==k_groups]`).

```{r cut-tree, echo=FALSE, message=FALSE, warning=FALSE}
# create groups
grp <- cutree(hclust1, k=k_groups)

# create data frame that identifies indicator species by group:
ivlab <- indval(x=ord.df, clustering = grp, numitr = 1000)

gr <- ivlab$maxcls 
iv <- ivlab$indcls 
pv <- ivlab$pval 

indvalsummary <- data.frame(group=gr, indval=iv, pvalue=pv) #, freq=fr)
indvalsummary <- indvalsummary[order(indvalsummary$group, -indvalsummary$indval),]
indvalsummary$Code <- rownames(indvalsummary)

# add in readable indicator names
indvalsummary <- left_join(x=indvalsummary,
                           y=select(indicator_descriptions, Indicator_code, Indicator),
                           by=c("Code"="Indicator_code")) %>%
  left_join(x=.,
            y=select(read.csv(data_file_paths(user)$species_list), SpeciesCode, ScientificName),
            by=c("Code"="SpeciesCode")) %>%
  mutate(Indicator = ifelse(test = is.na(Indicator), yes = ScientificName, no = Indicator))

if(opuntia_combined){
  indvalsummary$Indicator[which(indvalsummary$Code=="AH_OpuntiaCover")] <- indicator_descriptions$Indicator[which(indicator_descriptions$Indicator_code=="opuntia_combined")]
}

# add in relative abundance
relab <- as.data.frame(ivlab$relabu)
colnames(relab) <- paste0("RelativeAbund_", 1:ncol(relab))
relab$Code <- rownames(relab)
indvalsummary <- left_join(x=indvalsummary, y=relab, by="Code")

# add in relative frequency
relfr <- as.data.frame(ivlab$relfrq)
colnames(relfr) <- paste0("RelativeFreq_", 1:ncol(relfr)) 
relfr$Code <- rownames(relfr)
indvalsummary <- left_join(x=indvalsummary, y=relfr, by="Code")

# create a table for display where relative abundance and relative frequency are only displayed for the group for which the species is an indicator
indvalsummary_disp <- dplyr::select(indvalsummary, group, indval, pvalue, Indicator)
indvalsummary_disp$RelAbund <- NA
for(g in 1:k_groups){
 indvalsummary_disp[indvalsummary_disp$group==g, "RelAbund"] <- 
   indvalsummary[indvalsummary$group==g, paste0("RelativeAbund_", g)]
}

indvalsummary_disp$RelFreq <- NA
for(g in 1:k_groups){
 indvalsummary_disp[indvalsummary_disp$group==g, "RelFreq"] <- 
   indvalsummary[indvalsummary$group==g, paste0("RelativeFreq_", g)]
}

# order data frame by group, then significance; for output file
#indvalsummary_disp <- indvalsummary_disp[order(indvalsummary_disp$group, indvalsummary_disp$indval), ]
indvalsummary_disp <- indvalsummary_disp %>%
  group_by(group) %>%
  arrange(desc(indval), .by_group=T)

indvalsummary_disp_filt <- filter(indvalsummary_disp, pvalue<=0.05) %>%
  mutate(indval=round(indval, 2),
         pvalue=round(pvalue, 3),
         RelAbund=round(RelAbund, 2),
         RelFreq=round(RelFreq, 2)) %>%
  select(group, Indicator, indval, pvalue, RelAbund, RelFreq) %>%
  setNames(c("Group", "Indicator", "Indicator value", "p-value", "Relative abundance", "Relative frequency"))

kbl(x = indvalsummary_disp_filt, format = "html", row.names = F, align = "c",
    caption = "Indicator species for plant community groups. The indicator value of a species is a combined measure of fidelity (always present in the group) and specificity (exclusive to the group), where higher values indicate a stronger association with the given vegetation group. Relative abundance indicates specificity, where 1 means the species only occurs in plots of the given group. Relative frequency indicates fidelity, where 1 means the species always occurs in plots of the given group. Only species with a p-value less than 0.05 are displayed.") %>%
  kable_styling(bootstrap_options = "bordered") %>%
  row_spec(kable_input=., row = 0:nrow(indvalsummary_disp_filt), background = "lightsteelblue") %>%
  collapse_rows(columns = 1, valign = "top")

```

```{r dendrogram, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.cap="Cluster dendrogram for plant community groups. Boxes represent the groups identified by the indicator species analysis."}
# Cluster dendrogram plot
hclust1_dend <- as.dendrogram(hclust1)

# veg group color palette
pal_veg <- RColorBrewer::brewer.pal(n=k_groups, name = "Set1")

plot(hclust1_dend, leaflab= "none")
rect.dendrogram(hclust1_dend, 
                k=k_groups, 
                cluster = grp, 
                border = pal_veg[unique(cutree(hclust1_dend, k=k_groups)[order.dendrogram(hclust1_dend)])], 
                text = as.character(unique(cutree(hclust1_dend, k=k_groups)[order.dendrogram(hclust1_dend)])), 
                text_cex = 2, 
                lower_rect = -0.25,
                lwd =2)

#plot(ape::as.phylo(hclust1), tip.color = pal_veg9[grp], direction = "downwards")
```

```{r ordination-scree, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, results='hide', fig.keep='all', cache=FALSE, fig.cap="NMDS ordination scree plot comparing number of dimensions in the solution to stress."}
NMDS_scree(ord.df)
ord_dims <- 2 # change manually based on the plot results
```

```{r ordination, eval=TRUE, include=FALSE}

#run NMS ordination 
set.seed(1)
ord <- metaMDS(ord.df,
               k=ord_dims, # number of dimensions
               trymax = 30) 

```

Based on Figure \@ref(fig:ordination-scree), ordination stress is acceptable with `r ord_dims` dimensions, and stress reduction is minimal with additional dimensions. The stress for the ordination with `r ord_dims` dimensions is `r round(ord$stress, 3)`.

```{r ordination-plots, echo=FALSE, fig.height=5, fig.width=10, fig.cap="NMDS ordination of plot indicator data. Points represent plots; ellipses represent plant communities identified by the hierarchical clustering and indicator species analysis."}
# 2D plotting
plot_data_first$PlantCommunity <- as.factor(grp)

# manually update based on indicator species
group_labels <- c("1 - C4 perennial grassland",
                    "2 - Juniper woodland",
                    "3 - C3 perennial grasses and sagebrush",
                    "4 - Ephedra and biological soil crust")

par(mfrow=c(1,2))
# Axes 1x2
plot(ord, choices = c(1,2), type = "n", 
     xlim = c(-1.25, 1.25),
     ylim = c(-1, 1))
points(ord, choices = c(1,2), display = "sites",
       col=pal_veg[plot_data_first$PlantCommunity],
       pch = 21, cex = .6, bg=pal_veg[plot_data_first$PlantCommunity])
ordiellipse(ord, plot_data_first$PlantCommunity, col=pal_veg, lwd = 2, label = T)

# Legend
plot(ord, type = "n", axes=FALSE,
     display = "sites",
     col=pal_veg[plot_data_first$PlantCommunity],
     xlab = "",
     ylab = "")
legend(x="center",
       legend = group_labels,
       fill = pal_veg,
       title = "Plant communities")

par(mfrow=c(1,1))

```

## State and transition model {.tabset}

### CUSTOM DIAGRAM

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="STM"}
# TODO create custom diagram of states
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Legend"}
# TODO legend for custom diagram if needed
```

### STANDARD DIAGRAM

Click on state and transition labels to scroll to the respective text.

TODO create boxes for each state that link to the state description below

## State 1
## Reference State

TODO write state description

## Community 1.1
## Reference State

TODO write state description

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Annual production by plant type (representative values) or group (midpoint values)"}
# TODO create pie charts of community composition by functional group. May need to change "production" to "percent cover" to fit with our data
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of production (or percent cover) by functional type

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 5. Annual production by plant type")
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of ground cover by type

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 6. Ground cover")
```

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of func type percent cover at different heights

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 7. Canopy structure (% cover)")
```

### Additional community tables

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table of production (or percent cover) by species, with species linking to their USDA Plants profile

# knitr::kable(x = dom_plant_spp_tab, caption = "Table 8. Community 1.1 plant community composition")
```

# Interpretations

## Animal community

TODO write wildlife narrative

## Hydrological functions

TODO write hydrology narrative

## Recreational uses

TODO write recreation narrative

## Wood products

TODO write wood production narrative (probably "None")

## Other products

TODO grazing narrative could go here

# Supporting information

## Other references

## Contributors

+ Travis Nauman
+ Samuel Burch
+ Anna Knight
+ Mike Duniway

# Reference sheet

[Interpreting Indicators of Rangeland Health](https://wiki.landscapetoolbox.org/doku.php/field_methods:rangeland_health_assessment_i.e._indicators_of_rangeland_health) is a qualitative assessment protocol used to determine ecosystem condition based on benchmark characteristics described in the Reference Sheet. A suite of 17 (or more) indicators are typically considered in an assessment. The ecological site(s) representative of an assessment location must be known prior to applying the protocol and must be verified based on soils and climate. Current plant community cannot be used to identify the ecological site.

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# TODO create table with author, author contact info, date, approver, approved date, and "Composition (Indicators 10 and 12) based on" info for the IIRH reference sheet

# knitr::kable(x = dom_plant_spp_tab)
```

## Indicators

TODO fill in the 17 IIRH indicators for the reference state if available

TODO link to PDF of the reference sheet

# Print options

TODO look into options for printing from the RMarkdown HTML format


